Chemistry of the Orphan Stream I: Identifying Stream Members from Low-Resolution Spectroscopy

### 2012-09-15 2012-oss-mrs/draft-oss-mrs.tex
Adopted final distance of 23.5 +/- 1.0 kpc

\begin{abstract}
We present coordinates and properties of candidate K giant members in the Orphan Stream, which have been identified from low resolution spectroscopic data taken with AAOmega on the Anglo-Astronomical Telescope. With modest S/N and independent cuts in photometry, kinematics, gravity and metallicity we yield self-consistent, highly probable stream members. Our data indicates an overall stream metallicity of [Fe/H] $= -1.78 \pm 0.08$ which is more metal-rich than previously found and unbiased by spectral type. The metallicity dispersion in our stream sample is indicative that the Orphan Stream has a dSph origin. We find a revised stream distance of 24 kpc near the celestial equator, and our kinematic signature peaks at $V_{\textsc{GSR}} \sim ZZ$\,km s$^{-1}$. The observed velocity dispersion of our highest probability members is consistent with arising from the velocity uncertainties alone, indicating that, at least as regards the
line-of-sight, the Orphan Stream is kinematically quite cold. We highlight likely members for high resolution spectroscopic follow-up.
\end{abstract}

@@

The Orphan Stream was independently detected by both \citet{Grillmair;Dionatos_2006} and \citet{Belokurov;et-al_2006}, and is distinct from other substructures in the halo. The stream stretches over $60\,^\circ$ in the sky, has a low surface brightness and a narrow stream width of only $\sim2^\circ$ and as the name suggests, the parent object largely remains a mystery. The stream extends past the celestial equator \--- outside the SDSS footprint \--- but has not been detected in existing southern surveys \citep{Newberg;et-al_2010}. Whilst the parent system remains elusive, significant effort has been placed on associating the stream with known Milky Way satellites \citep{Zucker;et-al_2006, Fellhaur;et-al_2007,Jin;Lynden_Bell_2007,Sales;et-al_2008}. In contrast, there has been relatively limited observational work on the Orphan Stream itself other than the original discovery papers \citep{Grillmair;Dionatos_2006, Belokurov;et-al_2006, Belokurov;et-al_2007} and the work of \citet{Newberg;et-al_2010}. This is largely to be expected given the absence of multi-band photometry in the southern sky and the low total luminosity of the stream, making it difficult to reliably seperate Orphan Stream members from halo stars. Further understanding the extent of the Orphan Stream awaits the SkyMapper and Pan-STARRS photometric surveys \citep{Keller;et-al_2007, Hodapp;et-al_2004}.

@@

The data were reduced using the standard \textsc{2DFDR} reduction pipeline\footnote{http://www.aao.gov.au/2df/aaomega/aaomega\_2dfdr.html}. After flat-fielding, throughput calibration for each fibre was achieved using the intensity of skylines in each fibre. The median flux of dedicated sky fibres was taken for sky subtraction, and wavelength calibration was performed using arc lamp exposures between science frames. Three thirty minute science exposures were median combined to assist with cosmic ray removal. The median S/N obtained in the red arm for our fields is modest at 35 per pixel. With the presence of strong Ca \textsc{II} triplet lines in the red arm we are able to ascertain reliable radial velocities and reasonable estimates on overall metallicity \citep[][and references therein]{Starkenburg;et-al_2010}. Our spectral region also includes gravity-sensitive magnesium lines: Mg I at 8807\,{\AA}, and the Mg I\,b 3$p$-4$s$ triplet lines at $\sim$5178\,{\AA}. As we demonstrate in the next section, these lines are sufficient to discriminate dwarfs from giants even with weak signal.

@@

We have employed a combination of separate criterion to identify likely Orphan Stream members: kinematics, a giant/dwarf indication from Mg I lines, and consistent metallicities from isochrone fitting and the Ca\,\textsc{II} triplet lines. Each criterion is discussed here separately.

\subsection{Kinematics}
Radial velocities were measured by cross-correlating our normalised spectra against a K-giant synthetic template with a temperature of $4500$\,K, $\log{g}$ = 1 and [Fe/H] = $-1.5$ across the range $845 \leq \lambda \leq 870$\,nm. Heliocentric velocities were translated to the galactic rest frame by adopting the local standard of rest velocity as 220\,km s$^{-1}$ towards $(l, b) = (53^\circ, 25^\circ)$ \citep{Kerr;Lynden-Bell_1986, Mihalas;Binney_1981}\footnote{Where $V_{\textsc{gsr}} = V_{\textsc{helio}} + 220\sin{l}\cos{b} + 16.5\times[\sin{b}\sin{25} + \cos{b}\cos{25}\cos{(l - 53)}]$}. \citet{Newberg;et-al_2010} presented a comprehensive kinematic dataset on the Orphan Stream using BHB stars and employed a slightly different galactocentric transformation. This systematic difference equates to our galactocentric rest frame velocities being on average $\sim5$\,km s$^{-1}$ lower than those of \citet{Newberg;et-al_2010}.

% Besan\c{c}on heliocentric velocities have been translated to a galactocentric frame in the same way as our targets.
Figure \ref{fig:velocities} shows a histogram of our galactocentric velocities, compared to the predicted smooth line-of-sight velocity distribution for this region from the Besan\c{c}on model \citep{Robin;et-al_2003}. We have selected predicted stars from the Besan\c{c}on model using the same criterion outlined in \S\ref{sec:target-selection} after using the \citet{Jordi;et-al_2006} colour transformations. It is clear that our target selection has yielded mostly nearby thick disk dwarf stars with $V_{\textsc{gsr}} \approx -120$\,km s$^{-1}$. The expected kinematic signature for the Orphan Stream, as \citet{Newberg;et-al_2010} reported, is labelled in Figure \ref{fig:velocities}. There is no sharp kinematic peak representative of the Orphan Stream exists in our sample. From kinematics alone, our targets appears largely indistinguishable from a smooth halo distribution.

@@

In this region \citet{Newberg;et-al_2010} found the Orphan Stream to have $V_{\textsc{gsr}} = 110$ km s$^{-1}$ from BHB stars, approximately $105$\,km s$^{-1}$ on our scale. To isolate potential Orphan Stream members we have nominated a relatively wide kinematic selection region between $75 \leq V_{\textsc{gsr}} \leq 135$ km s$^{-1}$, which yields 25 Orphan Stream candidates. The typical uncertainty in our kinematics is $\pm{}3.5$\,km s$^{-1}$, although for our fainter targets this quickly increases above $10$ km s$^{-1}$.

@@

We have measured the equivalent widths of the gravity-sensitive Mg I line at 8807 \AA{} to distinguish dwarfs from giants \citep{Battaglia;Starkenburg_2012}. At a given temperature (or $g \-- r$) and metallicity, giant stars present narrower Mg I absorption lines than their dwarf contaminants. Given our target selection, our sample is likely to contain many more dwarfs than giants. In some cases no Mg I 8807 \AA{} line was apparent, so an upper limit was estimated based on the S/N. In these cases the candidate was considered a ``non-dwarf" because we cannot exclusively rule out a metal-poor sub-giant with this criterion alone. For these purposes we are only looking for a simple indication as to whether a star is likely a dwarf or not. Figure \ref{fig:ew-mg} illustrates the trend with EW$_{\lambda8807}$ against SDSS de-reddened\footnote{All magnitudes presented in this letter are de-reddened using the \citet{Schlegel;Finkbeiner;Davis_1998} dust maps.} $g \-- r$, illustrating the dominant upper dwarf branch we wish to exclude. The adopted dwarf/giant separation line is shown. Admittedly this selection criterion is not particularly strict, but we are allowing multiple inclusive cuts and taking the intersection of these criteria to identify Orphan Stream giants.

\begin{figure}[h]
	\includegraphics[width=\columnwidth]{./figures/ew-mg.pdf}
	\caption{De-reddened $g \-- r$ against the measured equivalent width of the Mg I transition at 8807 \AA{}. Dwarf contaminants occupy the more populous upper branch. Our separation line between dwarfs and giants is shown in green.}
	\label{fig:ew-mg}
\end{figure}

This analysis was also performed using the total equivalent width of the Mg\,\textsc{I}\,b triplet lines. Both analyses were entirely consistent with each other: essentially the same candidate selection was found using both techniques. However, given slightly poorer signal at the Mg\,\textsc{I}\,b triplet, we were forced to adopt many more upper limits than when using the 8807 {\AA} line. Because we classify all upper limits as being ``non-dwarfs" (i.e. potential OSS giants), we deduced a slightly larger candidate sample for the Mg\,\textsc{I}\,b analysis, which was primarily populated by upper limits. In conclusion, we found the Mg\,\textsc{I} line at 8807{\,\AA} appeared to be a more consistent dwarf discriminant given our weak S/N \--- particularly for our fainter stars \--- and thus we have used the 8807\,{\AA} Mg \textsc{I} selection throughout the rest of our analysis.

@@

\begin{figure}[h]
	\includegraphics[width=\columnwidth]{./figures/cmd.pdf}
	\caption{Color magnitude diagram showing our observed candidates (grey). Observations fulfilling kinematic and gravity cuts are colored by their metallicity, and those with upper limits for surface gravity are marked as triangles ($\triangledown$). Highly probable stream members with consistent line strength- and isochrone-derived metallicities are circled. Relevant 10 Gyr \citet{Girardi;et-al_2008} isochrones at $[\mbox{Fe/H}] = -1.5$ (dotted), $-2.0$ (dashed) at 21.4\,kpc \citep{Newberg;et-al_2010} are shown, as well as our best-fitting 10 Gyr isochrone of $[\mbox{Fe/H}] = -1.71$ at YY\,kpc (solid).}
	\label{fig:cmd}
\end{figure}

Given that our Orphan Stream giants cover a wide evolutionary range along the giant branch (Figure \ref{fig:cmd}), we are in a good position to revise the distance estimate to the stream. Given a 10 Gyr \citet{Girardi;et-al_2008} isochrone at $[\mbox{Fe/H}] = -1.71$, we find a best-fitting distance to the stream of YY\,kpc at ($l, $b) = ($250^\circ$, $50^\circ$). This isochrone is also shown in Figure \ref{fig:cmd}, and it is clear that the four most probable member stars lie almost precisely on this isochrone. Our derived distance is slightly farther than the $21.4 \pm 1.0$ kpc found by \citet{Newberg;et-al_2010}, but given the inherent uncertainties in determining distances we argue these measurements are largely in agreement.

@@

We have presented a detailed analysis to isolate individual Orphan Stream K giants from low-resolution spectroscopy using a combination of photometric, kinematic, gravity, and metallicity information. Although each individual criterion is likely to induce some level contamination, their intersection reveals 10 highly probable, self-consistent, Orphan Stream K giants.  We deduce a median stream metallicity of $[\mbox{Fe/H}] = -1.71$ and find an intrinsically wide metallicity spread of $\sigma = 0.56$\,dex, indicative of a dSph origin. Unlike other stellar tracers, K-type giants can exist at all metallicities, hence our derived metallicity spread is likely representative of the true stream metallicity distribution function. Recall that the metallicity determination was performed after kinematic and gravity cuts, and our most probable members lay perfectly on a 10 Gyr isochrone of $[\mbox{Fe/H}] = -1.71$. Our stream members indicate a revised distance to the stream of $YY$\,kpc at $(l, b) = (250\,^\circ$, $50\,^\circ)$, Z\,kpc slightly farther than that deduced by \citet{Newberg;et-al_2010}.

@@

Using a maximum-likelihood estimation we find the mean stream galactocentric velocity at $(l, b) = (250\,^\circ, 50\,^\circ)$ from five stars to be $V_{GSR} = ZZ.Z \pm Z.Z$ km s$^{-1}$ and the intrinsic dispersion to be $Z.Z \pm Z.Z$ km s$^{-1}$, i.e., comparable to our velocity resolution and consistent with zero. If we exclude five stars with weak signal \--- and hence large velocity uncertainties \--- this dispersion is found to be $Z \pm Z$\,km s$^{-1}$. These velocities indicate that the Orphan Stream is a very cold stream, and implies their motion is largely perpendicular to the line of sight.

>>>>

\begin{abstract}
We present coordinates and properties of candidate K giant members in the Orphan Stream, which have been identified from low resolution spectroscopic data taken with AAOmega on the Anglo-Astronomical Telescope. With modest S/N and independent cuts in photometry, kinematics, gravity and metallicity we yield self-consistent, highly probable stream members. Our data indicates an overall stream metallicity of [Fe/H] $= -1.71$ with an intrinsic dispersion of $\sigma = 0.56$\,dex which is more metal-rich than previously found and unbiased by spectral type. The metallicity dispersion in our stream sample is indicative that the Orphan Stream has a dSph origin. We find a revised stream distance of $23.5 \pm 1.0$\,kpc near the celestial equator, and our kinematic signature peaks at $V_{\textsc{GSR}} \sim ZZ$\,km s$^{-1}$. The observed velocity dispersion of our highest probability members is consistent with arising from the velocity uncertainties alone, indicating that, at least along the line-of-sight, the Orphan Stream is kinematically quite cold. We highlight likely members for high resolution spectroscopic follow-up.
\end{abstract}

@@

The Orphan Stream was independently detected by both \citet{Grillmair;Dionatos_2006} and \citet{Belokurov;et-al_2006}, and is distinct from other substructures in the halo. The stream stretches over $60\,^\circ$ in the sky, has a low surface brightness and a narrow stream width of only $\sim2\,^\circ$ and as the name suggests, the parent object largely remains a mystery. The stream extends past the celestial equator \--- outside the SDSS footprint \--- but has not been detected in existing southern surveys \citep{Newberg;et-al_2010}. Whilst the parent system remains elusive, significant effort has been placed on associating the stream with known Milky Way satellites \citep{Zucker;et-al_2006, Fellhaur;et-al_2007,Jin;Lynden_Bell_2007,Sales;et-al_2008}. In contrast, there has been relatively limited observational work on the Orphan Stream itself other than the original discovery papers \citep{Grillmair;Dionatos_2006, Belokurov;et-al_2006, Belokurov;et-al_2007} and the work of \citet{Newberg;et-al_2010}. This is largely to be expected given the absence of multi-band photometry in the southern sky and the low total luminosity of the stream, making it difficult to reliably seperate Orphan Stream members from halo stars. Further understanding the extent of the Orphan Stream awaits the SkyMapper and Pan-STARRS photometric surveys \citep{Keller;et-al_2007, Hodapp;et-al_2004}.

@@

The data were reduced using the standard \textsc{2DFDR} reduction pipeline\footnote{http://www.aao.gov.au/2df/aaomega/aaomega\_2dfdr.html}. After flat-fielding, throughput calibration for each fibre was achieved using the intensity of skylines in each fibre. The median flux of dedicated sky fibres was taken for sky subtraction, and wavelength calibration was performed using arc lamp exposures between science frames. Three thirty minute science exposures were median combined to assist with cosmic ray removal. The median S/N obtained in the red arm for our fields is modest at 35 per pixel. With the presence of strong Ca\,\textsc{II} triplet lines in the red arm we are able to ascertain reliable radial velocities and reasonable estimates on overall metallicity \citep[][and references therein]{Starkenburg;et-al_2010}. Our spectral region also includes gravity-sensitive magnesium lines: Mg\,\textsc{I} at 8807\,{\AA}, and the Mg\,\textsc{I}\,b 3$p$-4$s$ triplet lines at $\sim$5178\,{\AA}. As we demonstrate in the next section, these lines are sufficient to discriminate dwarfs from giants even with weak signal.

@@

We have employed a combination of separate criterion to identify likely Orphan Stream members: kinematics, a giant/dwarf indication from Mg\,\textsc{I} lines, and consistent metallicities from isochrone fitting and the Ca\,\textsc{II} triplet lines. Each criterion is discussed here separately.

\subsection{Kinematics}
Radial velocities were measured by cross-correlating our normalised spectra against a K-giant synthetic template with a temperature of 4500\,K, $\log{g}$ = 1 and $[\mbox{Fe/H}] = -1.5$ across the range $845 \leq \lambda \leq 870$\,nm. Heliocentric velocities were translated to the galactic rest frame by adopting the local standard of rest velocity as 220\,km s$^{-1}$ towards $(l, b) = (53\,^\circ, 25\,^\circ)$ \citep{Kerr;Lynden-Bell_1986, Mihalas;Binney_1981}\footnote{Where $V_{\textsc{gsr}} = V_{\textsc{helio}} + 220\sin{l}\cos{b} + 16.5\times[\sin{b}\sin{25} + \cos{b}\cos{25}\cos{(l - 53)}]$}. \citet{Newberg;et-al_2010} presented a comprehensive kinematic dataset on the Orphan Stream using BHB stars and employed a slightly different galactocentric transformation. This systematic difference equates to our galactocentric rest frame velocities being on average $\sim5$\,km s$^{-1}$ lower than those of \citet{Newberg;et-al_2010}.

Figure \ref{fig:velocities} shows a histogram of our galactocentric velocities, compared to the predicted smooth line-of-sight velocity distribution for this region from the Besan\c{c}on model \citep{Robin;et-al_2003}. We have selected predicted stars from the Besan\c{c}on model using the same criterion outlined in \S\ref{sec:target-selection} after using the \citet{Jordi;et-al_2006} colour transformations. It is clear that our target selection has yielded mostly nearby thick disk dwarf stars with $V_{\textsc{gsr}} \approx -120$\,km s$^{-1}$. The expected kinematic signature for the Orphan Stream, as \citet{Newberg;et-al_2010} reported, is labelled in Figure \ref{fig:velocities}. There is no sharp kinematic peak representative of the Orphan Stream exists in our sample. From kinematics alone, our targets appears largely indistinguishable from a smooth halo distribution.

@@

In this region \citet{Newberg;et-al_2010} found the Orphan Stream to have $V_{\textsc{gsr}} = 110$ km s$^{-1}$ from BHB stars, approximately $105$\,km s$^{-1}$ on our scale. To isolate potential Orphan Stream members we have nominated a relatively wide kinematic selection region between $75 \leq V_{\textsc{gsr}} \leq 135$ km s$^{-1}$, which yields 25 Orphan Stream candidates. The typical uncertainty in our kinematics is $\pm{}3.5$\,km s$^{-1}$, although for our fainter targets this quickly increases above 10\,km s$^{-1}$.

@@

We have measured the equivalent widths of the gravity-sensitive Mg\,\textsc{I} line at 8807 \AA{} to distinguish dwarfs from giants \citep{Battaglia;Starkenburg_2012}. At a given temperature (or $g \-- r$) and metallicity, giant stars present narrower Mg\,\textsc{I} absorption lines than their dwarf contaminants. Given our target selection, our sample is likely to contain many more dwarfs than giants. In some cases no Mg\,\textsc{I} 8807 \AA{} line was apparent, so an upper limit was estimated based on the S/N. In these cases the candidate was considered a ``non-dwarf" because we cannot exclusively rule out a metal-poor sub-giant with this criterion alone. For these purposes we are only looking for a simple indication as to whether a star is likely a dwarf or not. Figure \ref{fig:ew-mg} illustrates the trend with EW$_{\lambda8807}$ against SDSS de-reddened\footnote{All magnitudes presented in this letter are de-reddened using the \citet{Schlegel;Finkbeiner;Davis_1998} dust maps.} $g \-- r$, illustrating the dominant upper dwarf branch we wish to exclude. The adopted dwarf/giant separation line is shown. Admittedly this selection criterion is not particularly strict, but we are allowing multiple inclusive cuts and taking the intersection of these criteria to identify Orphan Stream giants.

\begin{figure}[h]
	\includegraphics[width=\columnwidth]{./figures/ew-mg.pdf}
	\caption{De-reddened $g \-- r$ against the measured equivalent width of the Mg\,\textsc{I} transition at 8807 \AA{}. Dwarf contaminants occupy the more populous upper branch. Our separation line between dwarfs and giants is shown in green.}
	\label{fig:ew-mg}
\end{figure}

This analysis was also performed using the total equivalent width of the Mg\,\textsc{I}\,b triplet lines. Both analyses were entirely consistent with each other: essentially the same candidate selection was found using both techniques. However, given slightly poorer signal at the Mg\,\textsc{I}\,b triplet, we were forced to adopt many more upper limits than when using the 8807\,{\AA} line. Because we classify all upper limits as being ``non-dwarfs" (i.e. potential OSS giants), we deduced a slightly larger candidate sample for the Mg\,\textsc{I}\,b analysis, which was primarily populated by upper limits. In conclusion, we found the Mg\,\textsc{I} line at 8807\,{\AA} appeared to be a more consistent dwarf discriminant given our weak S/N \--- particularly for our fainter stars \--- and thus we have used the 8807\,{\AA} Mg\,\textsc{I} selection throughout the rest of our analysis.

@@

\begin{figure}[h]
	\includegraphics[width=\columnwidth]{./figures/cmd.pdf}
	\caption{Color magnitude diagram showing our observed candidates (grey). Observations fulfilling kinematic and gravity cuts are colored by their metallicity, and those with upper limits for surface gravity are marked as triangles ($\triangledown$). Highly probable stream members with consistent line strength- and isochrone-derived metallicities are circled. Relevant 10 Gyr \citet{Girardi;et-al_2008} isochrones at $[\mbox{Fe/H}] = -1.5$ (dotted), $-2.0$ (dashed) at 21.4\,kpc \citep{Newberg;et-al_2010} are shown, as well as our best-fitting 10 Gyr isochrone of $[\mbox{Fe/H}] = -1.71$ at 23.5\,kpc (solid).}
	\label{fig:cmd}
\end{figure}

Given that our Orphan Stream giants cover a wide evolutionary range along the giant branch (Figure \ref{fig:cmd}), we are in a good position to revise the distance estimate to the stream. Given a 10\,Gyr \citet{Girardi;et-al_2008} isochrone at $[\mbox{Fe/H}] = -1.71$, we find a best-fitting distance to the stream of $23.5 \pm 1.0$\,kpc at ($l, $b) = ($250\,^\circ$, $50\,^\circ$). This isochrone is also shown in Figure \ref{fig:cmd}, and it is clear that the four most probable member stars lie almost precisely on this isochrone. Our derived distance is slightly farther than the $21.4 \pm 1.0$\,kpc found by \citet{Newberg;et-al_2010}, although given the difficulty in measuring distances these values are largely in agreement.

@@

We have presented a detailed analysis to isolate individual Orphan Stream K giants from low-resolution spectroscopy using a combination of photometric, kinematic, gravity, and metallicity information. Although each individual criterion is likely to induce some level contamination, their intersection reveals 10 highly probable, self-consistent, Orphan Stream K giants.  We deduce a median stream metallicity of $[\mbox{Fe/H}] = -1.71$ and find an intrinsically wide metallicity spread of $\sigma = 0.56$\,dex, indicative of a dSph origin. Unlike other stellar tracers, K-type giants can exist at all metallicities, hence our derived metallicity spread is likely representative of the true stream metallicity distribution function. Recall that the metallicity determination was performed after kinematic and gravity cuts, and our most probable members lay perfectly on a 10\,Gyr isochrone of $[\mbox{Fe/H}] = -1.71$. Our stream members indicate a revised distance to the stream of $23.5 \pm 1.0$\,kpc at $(l, b) = (250\,^\circ$, $50\,^\circ)$, slightly farther than that deduced by \citet{Newberg;et-al_2010}.

@@

Using a maximum-likelihood estimation we find the mean stream galactocentric velocity at $(l, b) = (250\,^\circ, 50\,^\circ)$ from five stars to be $V_{GSR} = ZZ.Z \pm Z.Z$\,km s$^{-1}$ and the intrinsic dispersion to be $Z.Z \pm Z.Z$\,km s$^{-1}$, i.e., comparable to our velocity resolution and consistent with zero. If we exclude five stars with weak signal \--- and hence large velocity uncertainties \--- this dispersion is found to be $Z \pm Z$\,km s$^{-1}$. These velocities indicate that the Orphan Stream is a very cold stream, and implies their motion is largely perpendicular to the line of sight.
